Virtualisierung eines Echtzeit-Betriebssystems zur Steuerung eines Roboters mit Schwerpunkt auf die Einhaltung der Echtzeit

### 2024-08-16 Masterarbeit_Robotik_Pamuk.tex
~\cite

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\columnwidth]{img/introduction/kernel_generic.png}
	\caption[Non-preemptible kernel]{Non-preemptible kernel \cite{WhatRealtimeLinuxa}}
	\label{fig:kernel_generic}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\columnwidth]{img/introduction/kernel_rt.png}
	\caption[Non-preemptible kernel]{Non-preemptible kernel \cite{WhatRealtimeLinuxa}}
	\label{fig:kernel_rt}
\end{figure}

@@

\section{Real-Time Operating Systems}
\noindent The RTOS structure is visualized in Figure~\ref{fig:rtos_structure}.The core component of an RTOS that enables real-time capabilities is the kernel. It is responsible for managing system resources, scheduling tasks, and ensuring deterministic behavior \cite{malallahComprehensiveStudyKernel2021}. It employs preemptive scheduling mechanisms to allow high-priority tasks to preempt lower-priority tasks, ensuring that time-critical tasks are not delayed. Additionally, RTOS kernels are designed to allocate and manage memory resources efficiently and minimize interrupt latency, which is crucial for real-time applications that require immediate response to external events~\cite{wangRealtimeEmbeddedSystems2017}.

@@

\bigskip \noindent \citeauthor{perneelRealtimeCapabilitiesStandard2015} \cite{perneelRealtimeCapabilitiesStandard2015} mention in their work that Linux was initially developed as a general purpose operating system without real-time applications in mind. They talk about the evolution of real-time behavior in the standard Linux kernel and state that it has recently gained popularity among the real-time community, largely due to its open-source nature and stability. The authors explain how to transform Linux into a real-time operating system by implementing the PREEMPT-RT patches. The paper also emphasizes some kernel configurations that ensure that the system's real-time behavior is preserved during runtime. Even though these modifications to the kernel enable soft real-time performance, the authors underline the fundamental rule in real-time software, that latency improvements have a negative impact on throughput and kernel performance. This is due to the added overhead of the \texttt{CONFIG\_PREEMPT\_RT} option. They concluded with their Linux build and a testing application that Linux can be a viable option for real-time applications, if it is correctly configured. Similarly, \citeauthor{adamRealTimePerformanceResponse2021} \cite{adamRealTimePerformanceResponse2021} applied the PREEMPT\_RT patch on both Raspberry Pi 3 and BeagleBone Black and achieved significantly lower latencies compared to standard Linux kernels.

\bigskip \noindent In the course of writing this thesis, the white paper "Real-Time Performance Tuning Best Practice Guidelines for KVM-Based Virtual Machines" \cite{RealTimePerformanceTuning2022} by Intel has been instrumental. Given that the working computer used for this work is equipped with an Intel processor, the guidelines presented in this document were particularly relevant and valuable for optimizing real-time performance in KVM-based virtual machines. Chapter \ref{cha:real-time_tuning} contains an extensive tuning process, which involves configurations spanning the BIOS, kernel, host OS, QEMU/KVM virtualization layer, and the Salamander 4 OS itself. The document concludes that the described tuning methods effectively improve real-time performance for KVM-based VMs, even under heavy workloads.

\bigskip \noindent \citeauthor{yoonRealTimePerformanceAnalysis} \cite{yoonRealTimePerformanceAnalysis} use some of these real-time tunings in their research, including CPU shielding, memory locking, and spinning nanosleep, for controlling humanoid robots equipped with around 60 servo motors and sensors. They implement EtherCAT for real-time communication of distributed devices whereas this thesis uses Varan. The authors highlight the importance of timely data processing in robots that collect data from their environment and respond accordingly through their sensors and actuators. The authors conclude with an experiment that their robotic system could achieve deterministic control of actuators and sensors even under heavy system load through the use of Linux with the real-time preemption patch. Their research has similarities to this master's thesis in terms of real-time tunings for robot control. However, this thesis includes also the virtualization aspect that is central.

\bigskip \noindent \citeauthor{sandstromVirtualizationTechnologiesEmbedded2013} \cite{sandstromVirtualizationTechnologiesEmbedded2013}, and \citeauthor{taccariEmbeddedRealTimeVirtualization2014} \cite{taccariEmbeddedRealTimeVirtualization2014} and  discuss and compare the current state-of-the-art in virtualization technologies with a focus on embedded real-time platforms. The latter emphasizes, that research on real-time systems is focusing on virtualization and multicore scheduling and underlines that flexibility and cost are key metrics. Both works acknowledge the great benefits of virtualization, that is, reducing overall hardware costs, since all non-critical activities and real-time control tasks can run on the same hardware. In addition, \citeauthor{javierperezHowRealTime2022} \cite{javierperezHowRealTime2022} state that hardware-based Programmable Logic Controllers (PLCs) are costly to maintain and lack the flexibility needed for modern, resource-intensive applications like Machine Learning (ML) or Artificial Intelligence (AI).

\bigskip \noindent Despite the promising advantages of virtualization, one must be aware of the potential drawbacks of it, too. \citeauthor{guStateoftheArtSurveyRealTime2012} \cite{guStateoftheArtSurveyRealTime2012} highlight issues like lock-holder preemption and task-grain scheduling. \citeauthor{garcia-vallsChallengesRealtimeVirtualization2014} \cite{garcia-vallsChallengesRealtimeVirtualization2014} identify challenges in supporting real-time applications in the cloud, including resource management, quality of service guarantees, temporal and spatial isolation, and network performance. \citeauthor{scordinoRealTimeVirtualizationIndustrial2020} \cite{scordinoRealTimeVirtualizationIndustrial2020} address the challenge of interference on shared hardware resources, which can degrade real-time performance.

\bigskip \noindent
\citeauthor{cinqueVirtualizingMixedCriticalitySystems2022} \cite{cinqueVirtualizingMixedCriticalitySystems2022}

\citeauthor{reghenzaniRealTimeLinuxKernel2020} \cite{reghenzaniRealTimeLinuxKernel2020}

\bigskip \noindent \citeauthor{maPerformanceTuningKVMbased} \cite{maPerformanceTuningKVMbased} \cite{junzhangPerformanceAnalysisKVMBased2010} describe in their work how KVM impacts the real-time performance, which has helped this work gain insight into the base overhead of virtualization. When a physical interrupt occurs while the guest RTOS is running, there are at least six stages that need to be traversed before the interrupt can be delivered to the guest RTOS. Once the interrupt is delivered to the guest RTOS, it incurs additional latency, called Interrupt Routine Time (IRT), since there is also the time between when an interrupt is recognized and the first instruction of the corresponding interrupt service routine (ISR) of the guest OS is started. These stages are the primary sources of latencies caused by KVM, they are briefly described in Figure \ref{fig:kvm_stages}.

@@

\bigskip \noindent \citeauthor{broskyShieldedProcessorsGuaranteeing2003} \cite{broskyShieldedProcessorsGuaranteeing2003} report in their work about the shielded CPU model for enhancing real-time performance in symmetric multiprocessor systems. By dedicating one or more CPUs to high-priority processes and interrupts, the goal is to provide deterministic execution of real-time applications and interrupt responses. The authors mention that there is no user interface for setting process CPU affinity in the standard Linux, while there is one for setting interrupt CPU affinity via the \texttt{/proc/irq/*/smp\_affinity} files. These files were used in this masther's thesis. According to the authors, the shielded CPU model is especially useful for tasks that require guaranteed interrupt response times, very fast interrupt responses and high-frequency, deterministic execution.

\bigskip \noindent \citeauthor{kiszkaLinuxRealTimeHypervisor} \cite{kiszkaLinuxRealTimeHypervisor} specifically focuses on Linux as a hypervisor and analyzes its real-time capabilities when using KVM and QEMU. Most importantly for this master's thesis, the author warns of starvation when the priorities of QEMU’s threads are raised and lifted into a real-time scheduling class. Starvation is when a process does not get the resources it needs for a long time because the resources are being allocated to other processes. A way to reduce this risk is by not using more virtual CPUs (VCPUs) than there are actual processor cores. Moreover, Linux has a feature that can limit how much CPU time all real-time tasks can use in the host system, which was an important information for the real-time tunings of this thesis later.

\bigskip \noindent \citeauthor{mckenneyRealTimeVs} \cite{mckenneyRealTimeVs} addresses the overheads associated with real-time Linux when using KVM and QEMU. The author lists several sources of overhead, including memory locking to avoid page-fault latencies, increased overhead of locking and interrupts due to more-aggressive preemption, threaded interrupts that permit long-running interrupt handlers to be preempted by high-priority real-time processes, real-time task scheduling that requires global scheduling, high-resolution timers for tens-of-microseconds accuracy and precision, and preemptible RCU which has slightly higher read-side overhead than Classic RCU. These information helped understand the trade-offs involved in real-time Linux systems.

\bigskip \noindent \citeauthor{deoliveiraDemystifyingRealTimeLinux} \cite{deoliveiraDemystifyingRealTimeLinux} critique the traditional \texttt{cyclictest} tool used for measuring scheduling latency, highlighting its limitations due to its black-box approach and lack of theoretical grounding. This is the main reason why this tool was not used to measure the scheduling latency in this work. Instead, the \texttt{latency} tool was used, as was determined in Chapter \ref{sec:application_context}.

>>>>

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\columnwidth]{img/introduction/kernel_generic.png}
	\caption[Non-preemptible kernel]{Non-preemptible kernel~\cite{WhatRealtimeLinuxa}}
	\label{fig:kernel_generic}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\columnwidth]{img/introduction/kernel_rt.png}
	\caption[Non-preemptible kernel]{Non-preemptible kernel~\cite{WhatRealtimeLinuxa}}
	\label{fig:kernel_rt}
\end{figure}

@@

\section{Real-Time Operating Systems}
\noindent The RTOS structure is visualized in Figure~\ref{fig:rtos_structure}.The core component of an RTOS that enables real-time capabilities is the kernel. It is responsible for managing system resources, scheduling tasks, and ensuring deterministic behavior~\cite{malallahComprehensiveStudyKernel2021}. It employs preemptive scheduling mechanisms to allow high-priority tasks to preempt lower-priority tasks, ensuring that time-critical tasks are not delayed. Additionally, RTOS kernels are designed to allocate and manage memory resources efficiently and minimize interrupt latency, which is crucial for real-time applications that require immediate response to external events~\cite{wangRealtimeEmbeddedSystems2017}.

@@

\bigskip \noindent~\citeauthor{perneelRealtimeCapabilitiesStandard2015}~\cite{perneelRealtimeCapabilitiesStandard2015} mention in their work that Linux was initially developed as a general purpose operating system without real-time applications in mind. They talk about the evolution of real-time behavior in the standard Linux kernel and state that it has recently gained popularity among the real-time community, largely due to its open-source nature and stability. The authors explain how to transform Linux into a real-time operating system by implementing the PREEMPT-RT patches. The paper also emphasizes some kernel configurations that ensure that the system's real-time behavior is preserved during runtime. Even though these modifications to the kernel enable soft real-time performance, the authors underline the fundamental rule in real-time software, that latency improvements have a negative impact on throughput and kernel performance. This is due to the added overhead of the \texttt{CONFIG\_PREEMPT\_RT} option. They concluded with their Linux build and a testing application that Linux can be a viable option for real-time applications, if it is correctly configured. Similarly,~\citeauthor{adamRealTimePerformanceResponse2021}~\cite{adamRealTimePerformanceResponse2021} applied the PREEMPT\_RT patch on both Raspberry Pi 3 and BeagleBone Black and achieved significantly lower latencies compared to standard Linux kernels.

\bigskip \noindent In the course of writing this thesis, the white paper "Real-Time Performance Tuning Best Practice Guidelines for KVM-Based Virtual Machines"~\cite{RealTimePerformanceTuning2022} by Intel has been instrumental. Given that the working computer used for this work is equipped with an Intel processor, the guidelines presented in this document were particularly relevant and valuable for optimizing real-time performance in KVM-based virtual machines. Chapter \ref{cha:real-time_tuning} contains an extensive tuning process, which involves configurations spanning the BIOS, kernel, host OS, QEMU/KVM virtualization layer, and the Salamander 4 OS itself. The document concludes that the described tuning methods effectively improve real-time performance for KVM-based VMs, even under heavy workloads.

\bigskip \noindent~\citeauthor{yoonRealTimePerformanceAnalysis}~\cite{yoonRealTimePerformanceAnalysis} use some of these real-time tunings in their research, including CPU shielding, memory locking, and spinning nanosleep, for controlling humanoid robots equipped with around 60 servo motors and sensors. They implement EtherCAT for real-time communication of distributed devices whereas this thesis uses Varan. The authors highlight the importance of timely data processing in robots that collect data from their environment and respond accordingly through their sensors and actuators. The authors conclude with an experiment that their robotic system could achieve deterministic control of actuators and sensors even under heavy system load through the use of Linux with the real-time preemption patch. Their research has similarities to this master's thesis in terms of real-time tunings for robot control. However, this thesis includes also the virtualization aspect that is central.

\bigskip \noindent~\citeauthor{sandstromVirtualizationTechnologiesEmbedded2013}~\cite{sandstromVirtualizationTechnologiesEmbedded2013}, and~\citeauthor{taccariEmbeddedRealTimeVirtualization2014}~\cite{taccariEmbeddedRealTimeVirtualization2014} and  discuss and compare the current state-of-the-art in virtualization technologies with a focus on embedded real-time platforms. The latter emphasizes, that research on real-time systems is focusing on virtualization and multicore scheduling and underlines that flexibility and cost are key metrics. Both works acknowledge the great benefits of virtualization, that is, reducing overall hardware costs, since all non-critical activities and real-time control tasks can run on the same hardware. In addition,~\citeauthor{javierperezHowRealTime2022}~\cite{javierperezHowRealTime2022} state that hardware-based Programmable Logic Controllers (PLCs) are costly to maintain and lack the flexibility needed for modern, resource-intensive applications like Machine Learning (ML) or Artificial Intelligence (AI).

\bigskip \noindent Despite the promising advantages of virtualization, one must be aware of the potential drawbacks of it, too.~\citeauthor{guStateoftheArtSurveyRealTime2012}~\cite{guStateoftheArtSurveyRealTime2012} highlight issues like lock-holder preemption and task-grain scheduling.~\citeauthor{garcia-vallsChallengesRealtimeVirtualization2014}~\cite{garcia-vallsChallengesRealtimeVirtualization2014} identify challenges in supporting real-time applications in the cloud, including resource management, quality of service guarantees, temporal and spatial isolation, and network performance.~\citeauthor{scordinoRealTimeVirtualizationIndustrial2020}~\cite{scordinoRealTimeVirtualizationIndustrial2020} address the challenge of interference on shared hardware resources, which can degrade real-time performance.

\bigskip \noindent
\citeauthor{cinqueVirtualizingMixedCriticalitySystems2022}~\cite{cinqueVirtualizingMixedCriticalitySystems2022}

\citeauthor{reghenzaniRealTimeLinuxKernel2020}~\cite{reghenzaniRealTimeLinuxKernel2020}

\bigskip \noindent~\citeauthor{maPerformanceTuningKVMbased}~\cite{maPerformanceTuningKVMbased}~\cite{junzhangPerformanceAnalysisKVMBased2010} describe in their work how KVM impacts the real-time performance, which has helped this work gain insight into the base overhead of virtualization. When a physical interrupt occurs while the guest RTOS is running, there are at least six stages that need to be traversed before the interrupt can be delivered to the guest RTOS. Once the interrupt is delivered to the guest RTOS, it incurs additional latency, called Interrupt Routine Time (IRT), since there is also the time between when an interrupt is recognized and the first instruction of the corresponding interrupt service routine (ISR) of the guest OS is started. These stages are the primary sources of latencies caused by KVM, they are briefly described in Figure \ref{fig:kvm_stages}.

@@

\bigskip \noindent~\citeauthor{broskyShieldedProcessorsGuaranteeing2003}~\cite{broskyShieldedProcessorsGuaranteeing2003} report in their work about the shielded CPU model for enhancing real-time performance in symmetric multiprocessor systems. By dedicating one or more CPUs to high-priority processes and interrupts, the goal is to provide deterministic execution of real-time applications and interrupt responses. The authors mention that there is no user interface for setting process CPU affinity in the standard Linux, while there is one for setting interrupt CPU affinity via the \texttt{/proc/irq/*/smp\_affinity} files. These files were used in this masther's thesis. According to the authors, the shielded CPU model is especially useful for tasks that require guaranteed interrupt response times, very fast interrupt responses and high-frequency, deterministic execution.

\bigskip \noindent~\citeauthor{kiszkaLinuxRealTimeHypervisor}~\cite{kiszkaLinuxRealTimeHypervisor} specifically focuses on Linux as a hypervisor and analyzes its real-time capabilities when using KVM and QEMU. Most importantly for this master's thesis, the author warns of starvation when the priorities of QEMU’s threads are raised and lifted into a real-time scheduling class. Starvation is when a process does not get the resources it needs for a long time because the resources are being allocated to other processes. A way to reduce this risk is by not using more virtual CPUs (VCPUs) than there are actual processor cores. Moreover, Linux has a feature that can limit how much CPU time all real-time tasks can use in the host system, which was an important information for the real-time tunings of this thesis later.

\bigskip \noindent~\citeauthor{mckenneyRealTimeVs}~\cite{mckenneyRealTimeVs} addresses the overheads associated with real-time Linux when using KVM and QEMU. The author lists several sources of overhead, including memory locking to avoid page-fault latencies, increased overhead of locking and interrupts due to more-aggressive preemption, threaded interrupts that permit long-running interrupt handlers to be preempted by high-priority real-time processes, real-time task scheduling that requires global scheduling, high-resolution timers for tens-of-microseconds accuracy and precision, and preemptible RCU which has slightly higher read-side overhead than Classic RCU. These information helped understand the trade-offs involved in real-time Linux systems.

\bigskip \noindent~\citeauthor{deoliveiraDemystifyingRealTimeLinux}~\cite{deoliveiraDemystifyingRealTimeLinux} critique the traditional \texttt{cyclictest} tool used for measuring scheduling latency, highlighting its limitations due to its black-box approach and lack of theoretical grounding. This is the main reason why this tool was not used to measure the scheduling latency in this work. Instead, the \texttt{latency} tool was used, as was determined in Chapter \ref{sec:application_context}.
